---
title: "nominate Codebook"
geometry: margin=.5in
output:
  html_document:
    df_print: paged
  pdf_document:
    fig_height: 2
    fig_width: 3
template: default
fontsize: 10pt
---

This data contain Kieth's Poole DW-NOMINATE measure of legislator ideology. Keith worked with various coauthors through the years, but Jeff Lewis now maintains the data, which you can find [here](https://voteview.com/data).

These data contain the first dimmensions of the DW-NOMINATE scores from Republican and Democrat members of the House from the 100th to the most recent Congress.

The data set is at the **member-Congress level**, so that each row of the data set represents one election in a particular Congress (e.g., Nancy Pelosi in the 110th Congress).

For details on the raw data, see the [codebook](https://voteview.com/articles/data_help_members) for the raw data.


## Load Data

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)

# load data
nominate_df <- read_rds("nominate.rds")

# quick look at data
glimpse(nominate_df)
```

## Variable Descriptions

### `congress`

- Coding: The number of the Congress
- Range: 100th to 114th
- Type: integer

### `state`

- Coding: The two-letter abbreviation of name of the member's state.
- Type: character

### `congressional_district`

- Coding: The number of the member's Congressional district. 
- Type: integer

### `party`

- Coding: The member's partisan affiliation. Members not affiliated with the Democratic or Republican Party are dropped from this data set.
- Type: character
- Values: 
  - `"Republican"`
  - `"Democrat"`

### `ideology`

- Coding: The first dimmension of the two-dimmenisonal DW-NOMINATE score. It is static over a member's career in Congress. Values near zero indicate a moderate. More negative values indicate a more liberal representative. More positive values indicate a more conservative representative. 
- Type: double

````{r}
# create a df with only the 114th congress
nominate114_df <- filter(nominate_df, congress == 114) 

# a histogram of the 114th congress
ggplot(nominate114_df, aes(x = ideology, fill = party)) + 
  geom_histogram(alpha = 0.5, position = "identity") + 
  scale_fill_manual(values = c("blue", "red")) + 
  labs(title = "Ideology in the 114th Congress")

# 5 most conservative members from 114th congress
top_n(nominate114_df, 5, ideology)

# 5 most liberal members from 114th congress
top_n(nominate114_df, 5, -ideology)

```